---
output: github_document
---

# <img src='man/figures/logo.svg' align="right" height="139" /> R package pomdp: Partially Observable Markov Decision Processes


```{r echo=FALSE}
options(digits = 2)
knitr::opts_chunk$set(tidy = TRUE, message = FALSE, fig.path = 'inst/README_files/')
```

```{r echo=FALSE, results = 'asis'}
pkg <- 'pomdp'

library(stringr)

  cat(str_interp("[![CRAN version](http://www.r-pkg.org/badges/version/${pkg})](https://CRAN.R-project.org/package=${pkg})\n"))
  cat(str_interp("[![stream r-universe status](https://mhahsler.r-universe.dev/badges/${pkg})](https://mhahsler.r-universe.dev/ui#package:${pkg})\n"))
  cat(str_interp("[![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/grand-total/${pkg})](https://CRAN.R-project.org/package=${pkg})\n"))
```

Provides the infrastructure to define and analyze the solutions of optimal control problems formulated as Partially Observable Markov Decision Processes (POMDP). The package uses the solvers from [pomdp-solve](http://www.pomdp.org/code/) (Cassandra, 2015) available in the R package [**pomdpSolve**](https://github.com/mhahsler/pomdpSolve) to solve POMDPs using
a variety of algorithms.

The package provides the following algorithms:

* Exact value iteration
  - __Enumeration algorithm__ (Sondik 1971, Mohan 1982).
  - __Two pass algorithm__ (Sondik 1971).
  - __Witness algorithm__ (Littman, Cassandra, Kaelbling 1996).
  - __Incremental pruning algorithm__ (Zhang and Liu 1996, Cassandra et al 1997).

* Approximate value iteration
  - __Finite grid algorithm__ (Cassandra 2015), a variation of point-based value iteration to solve larger POMDPs (__PBVI__; see Pineau 2003) without dynamic belief set expansion.
  - __SARSOP__ (Kurniawati, Hsu and Lee 2008), point-based algorithm that approximates optimally reachable belief spaces for infinite-horizon problems (via package [sarsop](https://github.com/boettiger-lab/sarsop)).

## Installation

__Stable CRAN version:__ install from within R with
```{r eval=FALSE}
install.packages("pomdp")
```

__Current development version:__ Install from [r-universe.](https://mhahsler.r-universe.dev/ui#package:pomdp)

## Usage

Solving the simple infinite-horizon Tiger problem.
```{r}
library("pomdp")
data("Tiger")
Tiger
```

```{r}
sol <- solve_POMDP(model = Tiger)
sol
```

```{r value_function, fig.height=4}
plot_value_function(sol, ylim = c(0, 20))
```

```{r policy_graph, fig.height=7}
plot_policy_graph(sol)
```

## References

* Cassandra, A. (2015). pomdp-solve: POMDP Solver Software, http://www.pomdp.org.
* Sondik, E. (1971). The Optimal Control of Partially Observable Markov Processes. Ph.D. Dissertation, Stanford University.
* Cassandra, A., Littman M.L., Zhang L. (1997). Incremental Pruning: A Simple, Fast, Exact Algorithm for Partially Observable Markov Decision Processes. UAI'97: Proceedings of the Thirteenth conference on Uncertainty in artificial intelligence, August 1997, pp. 54-61.
* Monahan, G. E. (1982). A survey of partially observable Markov decision processes: Theory, models, and algorithms. Management Science 28(1):1-16.
* Littman, M. L.; Cassandra, A. R.; and Kaelbling, L. P. (1996). Efficient dynamic-programming updates in partially observable Markov decision processes. Technical Report CS-95-19, Brown University, Providence, RI.
* Zhang, N. L., and Liu, W. (1996). Planning in stochastic domains: Problem characteristics and approximation. Technical Report HKUST-CS96-31, Department of Computer Science, Hong Kong University of Science and Technology.
* Pineau, J., Gordon, G.J., Thrun, S.B. (2003). Point-based value iteration: an anytime algorithm for POMDPs. IJCAI'03: Proceedings of the 18th international joint conference on Artificial Intelligence. Pages 1025-1030.
* Kurniawati, H., Hsu, D., and Lee, W.S. (2008). SARSOP: Efficient point-based POMDP planning by approximating optimally reachable belief spaces. In Proc. Robotics: Science and Systems.

## Acknowledgments

Development of this package was supported in part by 
National Institute of Standards and Technology (NIST) under grant number 
[60NANB17D180](https://www.nist.gov/ctl/pscr/safe-net-integrated-connected-vehicle-computing-platform).
