---
title: "Solving Markov Decision Processes with pomdp"
author: "Michael Hahsler"
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Solving Markov Decision Processes with pomdp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pomdp)
```

# Introduction

A Markov decision process (MDP) describes sequential decision making when the
state of the environment is completely observable. At every epoch, the agent
observes the current state, chooses an action, receives a reward, and moves to a
new state according to the transition probabilities. The objective is to find a
policy that maximizes the expected sum of future rewards.

A finite-state MDP is described by states $S$, actions $A$, transition
probabilities $T(s' \mid s,a)$, rewards $R(s,a,s')$, and a discount factor
$\gamma$. The package **pomdp** represents these components with an `MDP`
object and solves the model with `solve_MDP()`. Both dynamic programming and
temporal-difference control methods are available [@Sutton1998].

The basic workflow is:

1. Define a model with `MDP()` or load one of the included examples.
2. Find a policy with `solve_MDP()`.
3. Inspect, visualize, evaluate, or simulate the policy.

# An Example MDP

We use the included `Maze` model. It is a gridworld with one start state, two
terminal states, and one wall. The agent can move up, right, down, or left.

```{r}
data(Maze)
Maze
```

An MDP is stored as a list. Its most important components are the states,
actions, transition model, reward model, discount factor, horizon, and start
distribution.

```{r}
Maze$states
Maze$actions
Maze$discount
Maze$horizon
Maze$start
```

Use `transition_matrix()` and `reward_matrix()` to inspect normalized model
components. For example, the following matrix contains the transition
probabilities for moving up.

```{r}
transition_matrix(Maze, action = "up")
```

Models can be created directly with `MDP()`. Transition probabilities can be
specified as matrices, data frames built with `T_()`, or functions. Rewards can
similarly be specified as matrices, data frames built with `R_()`, or
functions. See `?MDP` for complete examples and the supported shortcuts.

# Solving an Infinite-Horizon MDP

`solve_MDP()` provides two model-based dynamic programming methods and three
model-free temporal-difference methods:

- `"value_iteration"` applies Bellman optimality backups until the value
  function converges.
- `"policy_iteration"` alternates approximate policy evaluation and greedy
  policy improvement.
- `"q_learning"` learns an off-policy action-value function.
- `"sarsa"` learns an action-value function on-policy.
- `"expected_sarsa"` uses the expected value under the current policy in its
  update.

Value iteration is the default. It is a good starting point when the complete
transition and reward models are available.

```{r}
maze_solved <- solve_MDP(Maze, method = "value_iteration")
maze_solved
```

The returned object contains the original model and a `solution` component.
For an infinite-horizon dynamic programming solution, the latter records the
method, convergence status, final value-function change, number of iterations,
and policy.

```{r}
maze_solved$solution[c("method", "converged", "delta", "iterations")]
```

`policy()` extracts the solution as one row per state. The column `U` is the
expected discounted return from that state, and `action` is the action selected
by the policy.

```{r}
policy(maze_solved)
```

The action-value function $Q(s,a)$ gives the expected return from taking action
$a$ in state $s$ and then following the policy. It is useful for comparing the
available actions in each state.

```{r}
round(q_values_MDP(maze_solved), 2)
```

The value function and the gridworld policy can also be visualized.

```{r, fig.width=6, fig.height=4}
plot_value_function(maze_solved)
```

```{r, fig.width=6, fig.height=4}
gridworld_plot_policy(maze_solved)
```

# Choosing Dynamic Programming Parameters

For value iteration, `error` controls the maximum policy loss used in the
convergence criterion, while `N_max` limits the number of iterations. An
initial value function can be supplied with `U`.

```{r}
maze_precise <- solve_MDP(
  Maze,
  method = "value_iteration",
  error = 1e-5,
  N_max = 1000
)
maze_precise$solution[c("converged", "delta", "iterations")]
```

Modified policy iteration starts with a random policy. In each iteration it
uses `k_backups` Bellman backups to approximately evaluate the current policy,
then replaces it with a greedy policy. It stops when the policy no longer
changes.

```{r}
maze_policy_iteration <- solve_MDP(
  Maze,
  method = "policy_iteration",
  k_backups = 10
)
maze_policy_iteration$solution[c("converged", "iterations")]
policy(maze_policy_iteration)
```

The two dynamic programming methods may make different choices when actions
have the same value, but their policies have the same optimal value function.

# Finite-Horizon Problems

For a finite horizon, the best action can depend on how many epochs remain.
Specify the number of epochs with `horizon`. Finite-horizon solving currently
uses value iteration and returns one policy per epoch.

```{r}
maze_finite <- solve_MDP(
  Maze,
  method = "value_iteration",
  horizon = 3
)
policy(maze_finite)
```

Individual epochs can be extracted or plotted separately.

```{r}
finite_policy <- policy(maze_finite, drop = FALSE)
finite_policy[[1]]
finite_policy[[3]]
```

```{r, fig.width=6, fig.height=4}
gridworld_plot_policy(maze_finite, epoch = 1)
```

# Learning from Simulated Experience

The temporal-difference methods use the MDP model only to simulate interaction
with the environment. They are useful for experimenting with reinforcement
learning. Their main controls are the number of episodes `N`, learning rate
`alpha`, exploration rate `epsilon`, and maximum episode length `horizon`.
Models with no absorbing state require a finite `horizon`.

The algorithms involve random exploration, so set a seed when reproducibility
is important.

```{r}
set.seed(1234)
maze_q_learning <- solve_MDP(
  Maze,
  method = "q_learning",
  N = 1000,
  alpha = 0.1,
  epsilon = 0.1
)
policy(maze_q_learning)
```

Q-learning is off-policy: it explores with an epsilon-greedy behavior policy
while learning a greedy target policy. Sarsa and expected Sarsa instead learn
from the current behavior policy. Because these methods estimate a solution
from sampled experience, their learned values and policies can vary between
runs and generally require parameter tuning.

# Evaluating and Simulating a Policy

`MDP_policy_evaluation()` applies Bellman backups to estimate the value of a
given policy. This is useful for comparing an optimal policy with a manually
specified or random policy.

```{r}
set.seed(1234)
random_policy <- random_MDP_policy(Maze)

rbind(
  random = MDP_policy_evaluation(random_policy, Maze),
  optimal = MDP_policy_evaluation(policy(maze_solved), Maze)
)
```

`simulate_MDP()` follows a solved policy through the environment. It reports
the average discounted reward and visitation counts, and can optionally return
complete trajectories.

```{r}
set.seed(1234)
simulation <- simulate_MDP(
  maze_solved,
  n = 1000,
  horizon = 20,
  return_trajectories = TRUE
)
simulation$avg_reward
head(simulation$trajectories)
```

Simulation is especially helpful for checking how a policy behaves from a
particular start distribution, estimating its realized reward, and diagnosing
states or actions that occur frequently.

# Next Steps

The `Maze`, `Cliff_walking`, `Windy_gridworld`, and `DynaMaze` data sets provide
additional MDP examples. See `vignette("gridworlds", package = "pomdp")` for
tools that build and visualize gridworld MDPs, and `?solve_MDP` for all solver
parameters.
